### 2020-07-06 paper.Rmd
work on k-means formalization

N <- 14
K <- 2
partitions <- generate_partitions(N, K)
length(partitions) # number of possible partitions
features <- rnorm(N)

@@

The plot illustrates all possible ways to partition a data set of $N = 
14$ into two parts. Red point: maximum similarity in means; green point 
= maximum similarity in variance. Blue point: Almost no difference to 
red point wrt similarity in means, but variance is much more similar. 
This blue partition is certainly preferable to the red one when maximal 
overall similarity is required; the red one just optimizes similarity in 
means (which is what maximizing the k-means objective does).

@@

The popular k-means clustering method aims at minimizing the 
within-group variance, i.e., the sum of the squared Euclidean distances 
between each data point and its cluster center:

More specifically:

$$
\mathit{Var}_{\mathrm{within}} = \sum\limits_{j=1}^{K} 
\sum\limits_{x_i \in c_j} \vert \vert x_i - \mu_j \vert \vert^2
$$

The squared distance between an element $x$ and the average of all 
elements $\overline{\mathrm{\boldsymbol{x}}}$ is:

>>>>

N <- 14
K <- 2
partitions <- generate_partitions(N, K)
features <- rnorm(N)

@@

The plot illustrates all possible ways to partition a data set of $N = 
14$ into two parts (there are `r length(partitions)` possible 
partitions). Red point: maximum similarity in means; green point 
= maximum similarity in variance. Blue point: Almost no difference to 
red point wrt similarity in means, but variance is much more similar. 
This blue partition is certainly preferable to the red one when maximal 
overall similarity is required; the red one just optimizes similarity in 
means (which is what maximizing the k-means objective does).

@@

In my treatment of k-means clustering, I adopt the notation Steinley 
(2006) provided in his thorough synthesis of the k-means literature. In 
clustering---or anticlustering---, an item pool $X$ = $\{x_1, \ldots, 
x_N\}$ has to be partitioned into $K$ groups $\{C_1, \ldots, C_K\}$. We 
will refer to the groups $C_j$, $j = 1, \ldots, K$, as anticlusters, 
clusters, or groups, and to the collection of anticlusters $C$ as the 
partitioning of the data set. We assume that each element $x_i$ ($i = 1, 
\ldots, N$), is a vector of length $P$, and each entry describes one of 
its numeric measurements. Generally hence, $X$ can be interpreted as an 
$N \times M$ matrix where each row represents an data point $x_i$ (e.g., 
representing a stimulus or a person) and each column is a numeric 
attribute (e.g., the number of syllables in a word, or the 
attractiveness rating of a face).

K-Means anticlustering aims to maximize the error sum of squares (SSE), 
also known as the within-group variance, which is the sum of the 
squared Euclidean distances between each data point and its cluster 
centroid. The cluster centroid is given by averaging the values on each 
variable over the elements within that cluster. The centroid of the 
*j*th variable in the *k*th cluster, consisting of $n_k$ elements, is 
given as:

$$
\overline{x}_j ^{(k)} = \frac{1}{n_k} \sum\limits {i \in C_k} x_{ij}
$$

The k-means objective function, the error sum of squares, is then given 
as:

$$
\mathit{SSE} = 
  \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (x_{ij} - \overline{x}_j ^{(k)})^2
$$

---

**This is old, adjust to Steinley (2006):**

The squared distance between an element $x = (x_1, \ldots, x_M)$ and the 
average of all elements $\overline{\mathrm{\boldsymbol{x}}}$ is:
